# Data visualization with ggplot2

```{r, echo=FALSE, message=FALSE, eval=FALSE}
# Set eval=TRUE to hide all results and figures.
# This sets defaults. Can change this manually in individual chunks.
# Must load knitr so opts_chunk is in search path.
library(knitr)
opts_chunk$set(results="hide", fig.show="hide", fig.keep="none")
```

```{r, echo=FALSE, message=FALSE, eval=TRUE}
library(knitr)
opts_chunk$set(message=FALSE)
```


## ggplot2 

No we're going to introduce a data visualization package called **ggplot2**.
This package is great for producing publication quality graphics, but the syntax
used in plotting can be a bit tricky.

First, recall how to install and load packages. Install the package if you haven't already:

```{r installggplot2, eval=FALSE}
# Only need to do this once
install.packages("ggplot2")
```

Then load it:

```{r loadggplot2, message=FALSE}
library(ggplot2)
```


Let's retun to the iris dataset, using the dataset we created that includes habitat type.

```{r look-at-iris}
head( iris_full )
tail( iris_full )
```

We would like to plot Sepal.Length versus Petal.Length. We'll first do this using
**base R** functions.

```{r plot1}
plot( x = iris_full$Sepal.Length, y = iris_full$Petal.Length )
```

Note, we could save a little typing by using the `with` function. This 
sets up a temporary environment where all of the variables (columns) 
in my dataset are defined individually.

```{r plot1-with}
with( data = iris_full, plot( x = Sepal.Length, y = Petal.Length ) )
```

OK, let's fire up ggplot.

```{r plot1-ggplot}
ggplot() + # First make an empty ggplot canvas. Notice the trailing plus sign
  geom_point( data = iris_full, aes( x = Sepal.Length, y = Petal.Length ) )
```

Let's break down that call to introduce a few key things about ggplot

* ggplot: the initial canvas we're working on
* geom: geometric objects (i.e. the type of plot - histogram, points, line, etc) 
* aes: aesthetic mapping 

**THAT SEEMS SO COMPLICATED!**

It's true, for simple plots, ggplot can be much more complicated
than simply using the base functions. But the power of ggplot lies
in the ability to lay several geometries (geoms) over each other.
Also, each geometry has a rich set of options. For example,
let's say I want to create the plot we just made, but have 
each species represented by a different color.

```{r plot2}
ggplot() + # First make an empty ggplot canvas. Notice the trailing plus sign
  geom_point( data = iris_full, aes( x = Sepal.Length, y = Petal.Length, colour = Species ) )
```

**facets** - a way to seperate data into different subplots

Let's say we wanted different plots for each species. We can do that in ggplot using
`facets`

```{r plot2}
ggplot( data = iris_full, aes( x = Sepal.Length, y = Petal.Length ) ) + 
  geom_point() +
  facet_grid( Species ~ . )
```








## Further **ggplot2** resources

* <http://docs.ggplot2.org/current/>: The official **ggplot2** documentation.
* <http://www.amazon.com/o/ASIN/0387981403/ref=nosim/gettgenedone-20>: The **ggplot2** book, by the developer, Hadley Wickham.
* <https://groups.google.com/forum/#!forum/ggplot2>: The **ggplot2** Google Group (mailing list, discussion forum).
* <https://github.com/swcarpentry/bc/tree/master/intermediate/r/data-visualization>: Intermediate Software Carpentry lesson on data visualization with **ggplot2**.
* <http://learnr.wordpress.com/>: A blog with a good number of posts describing how to reproduce various kind of plots using **ggplot2**.
* <http://stackoverflow.com/questions/tagged/ggplot2>: Thousands of questions and answers tagged with "ggplot2" on Stack Overflow, a programming Q&A site.
