---
title: "Working with dplyr"
output: html_document
---

We just briefly covered a few basics of working with relational databases. 
Now, we're going to go back to working with in R, and work with a package that
borrows much of it's syntax from SQL type languages. This package is called
`dplyr`

The first step is to install the `dplyr` package. We'll use the command
`install.packages` to do this. The first time you use `install.pacakges` you 
may see a prompt asking you to choose a "mirror site" to download from. 
You can select any site, but generally the closer the better.

```
install.packages( "dplyr", dependencies = TRUE )
```

Now that we have installed `dplyr`, let's load it into our R session.
We'll use the `library` command to do this.

```{r}
library( dplyr )
```

`dplyr` is a package with functions to help us do many of the common 
tasks we often want to do with our data. It is well documented, with
an excellent 
[vignette](http://cran.rstudio.com/web/packages/dplyr/vignettes/introduction.html) 
to get you up and running.

We're going to work with a dataset that comes built in to R, commonly 
called the
[iris dataset](https://stat.ethz.ch/R-manual/R-patched/library/datasets/html/iris.html).
It is also sometimes called Fisher's Iris dataset (but should more appropriately
be called Anderson's Iris dataset). Because it comes pre-packaged with R, we 
can load it into our environment using the `data` function.

```{r}
data( iris )
```

Let's have a look at this dataset

```{r}
head( iris )
tail( iris )
```

The dataset contains measurements of four characteristics of three
different species of Iris (plants!). 

## dplyr::select

Recall our lesson on selecting data in sql. We may want to do
something similar here - let's say we want to select from
the Iris dataset only the Species names and the Petal.Length 
measurement.

```{r}
select( iris, Species, Petal.Length )
```

## dplyr::arrange

Now let's say we want to order our data by one of our measurement variables. 
In SQL we used `ORDER BY`. With `dplyr` we're going to use a function
names `arrange` Note that to keep things cleaner I'm only going to 
take the head and tail of these new datasets.

```{r}
head( arrange( iris, Sepal.Length ) )
tail( arrange( iris, Sepal.Length ) )
```

## dplyr::filter

Let's filter our data set to we are only looking at specimens with
Petal.Length greater than 5. This is similar to SQL FILTER.

```{r}
head( filter( iris, Petal.Length > 5 ) )
```

## dplyr::mutate

What if we want to create new variables using the variables
in our current dataset?

```{r}
head( mutate( iris, Petal.Area = Petal.Width * Petal.Length ) )
tail( mutate( iris, Petal.Area = Petal.Width * Petal.Length ) )
```

## Connecting functions

Often we would like to carryout several of the above operations 
on our data. One way to do this would be to perform each operation
separately, saving the output, then feeding the output from one operation
to the next operation. For example, let's say we want to filter by 
Petal.Width less than 2, then estimate Petal.Area. We could do ...

```{r}
temp1 <- filter( iris, Petal.Width < 2 )
temp2 <- mutate( temp1, Petal.Area = Petal.Width * Petal.Length )
head( temp2 )
```

However, just like we can "pipe" output from one command into
a new command in the Unix Shell environment, there is a function
that allows us to do the same thing in R.

```{r}
temp3 <- 
  filter( iris, Petal.Width < 2 ) %>%
  mutate( Petal.Area = Petal.Width * Petal.Length )

head( temp3 )
```

And for one more level of piping we can do the following:

```{r}
temp4 <-
  iris %>%
  filter( Petal.Width < 2 ) %>%
  mutate( Petal.Area = Petal.Width * Petal.Length )

head( temp4 )
```

Note that we no longer need to have the argument defining the 
dataset we are interested in working with when we are piping to
a function. The data are assumed to be the output of the previous
command.

***

# Challange

Use the functions we just learned, along with the pipe function (%>%)
to: 

1. Filter out iris with Petal.Length greater than 2
2. Calculate Petal.Area
3. Sort the dataset by Petal.Area

***

## dplyr::summarise

The `summarise` function allows us to calculate summaries of our 
data, such as mean, min, etc. Let's calculate the mean for one 
of our measurement variables.

```{r}
iris %>%
  summarise( Sepal.Length.Mean = mean( Sepal.Length ) )

```

This is clearly more complicated than simply executing 
`mean( iris[,1] )` or `mean( iris$Sepal.Length )`, but the utility
of `summarise` will be evident in just a minute.

## dplyr::group_by

What if we want mean values of our measurements for each of the 
three species? This is where the `group_by` function comes in.

```{r}
iris %>%
  group_by( Species ) 
```

Nothing much seems to have happened, but note that the output
now includes a line starting with "Groups: ". This means 
that any `summarise` or `mutate` operations we use will now
be applied across groups separately. For example:

```{r}
iris %>%
  group_by( Species ) %>%
  summarise( Sepal.Len.Mean = mean( Sepal.Length ) )
```

We can ask for multiple `summarise`s at the same time.

```{r}
iris %>%
  group_by( Species ) %>%
  summarise( Sepal.Len.Mean = mean( Sepal.Length ),
             Sepal.Wid.Mean = mean( Sepal.Width ) )
```

***
# Challange

Write a script that returns the mean, max, and min values,
by species, for Sepal.Width.

***

## Merging datasets

Another common task, and one related to dealing with database tables, is to 
merge related datasets. For example, let's say we had a dataset describing
the habitat preferences for each of the three species of Iris.

```{r}
iris_habitat <- data.frame( Species = c( "setosa", "versicolor", "virginica" ),
                            Habitat = c( "forest", "wetland", "meadow" ) )
```

How can we merge these data? We'll look at two ways.

First - use the `merge` command.

```{r}
iris_full <- merge( x = iris, y = iris_habitat, by = "Species" )
head( iris_full )
tail( iris_full )
```

**WARNING. This next section is here primarily to provide you with a reference
for you might begin to think about merging data tables in the future. Do not
be concerned if this is over your head.**

Second (more advanced) - this method takes advantage of a variable type called "data.tables". 
Manipulations of these variable types are very similar to SQL. You may have to 
install this package first.

```{r}
library( data.table )
iris_tbl <- data.table( iris )
iris_habitat_tbl <- data.table( iris_habitat )

setkey( iris_tbl, Species )
setkey( iris_habitat_tbl, Species )

iris_tbl[ iris_habitat_tbl ]
```

Look up the help file for `setkey`

## Bringing it all back together with databases

We can work directly with SQL databases within R, primarily using
`dplyr` and `data.table`. 

# Working with sql databases using the dplyr package
# http://cran.r-project.org/web/packages/dplyr/vignettes/databases.html

```{r}
gen_db <- src_sqlite( "/Users/mlammens/Google Drive/Professional/Software-Carpentry/2015-03-06-nyas/lessons/sql/survey.db" )

gen_db

person <- tbl(gen_db, from = "Person" )

person %>%
  select( personal, family )

## rather than select, can also de-select
person %>%
  select( -personal )



```


***

# Challange

Merge the Person and Survey. Hint, you can only use `merge` by 
first converting these objects into `data.frames` (rather than
`data.tables`), then allow the 